---
title: "ToothGrowth Data Exploratory Analysis"
author: "Alexander Tuzhikov"
date: "September 14, 2015"
output:
  pdf_document:
    fig_height: 5
    fig_width: 8
    highlight: haddock
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 1
  html_document:
    fig_caption: yes
    fig_height: 9
    fig_width: 12
    highlight: haddock
    keep_md: yes
    number_sections: yes
    toc: yes
mainfont: Ubuntu
documentclass: article
---
#Synopsis

We are now moving to the part 2 of the task. Below we will exlore the `ToothGrowth` dataset from R `datasets` package. The headers below will correspond to the tasks. The data comes from the study [The Effect of Vitamin C on Tooth Growth in Guinea Pigs"][1]. The data is described as: *The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).*

#Load the ToothGrowth data and perform some basic exploratory data analyses

```{r libraries and data, message=FALSE, warning=FALSE}
library(datasets)
library(gridExtra)
library(ggplot2)
library(dplyr)
data("ToothGrowth")
```

#Provide a basic summary of the data
Well, the data is presented as a `data.frame` of `r nrow(ToothGrowth)` samples in `r ncol(ToothGrowth)` rows: `r colnames(ToothGrowth)`; 

* **len** is the tooth length (microns)
* **supp** is the supplement type (VC or OJ)
* **dose** is the dose (milligrams)

```{r summary 1.1, echo=FALSE}                         
ToothGrowth %>% mutate(dose=factor(dose)) %>% summary()
```
```{r summary 1.2, echo=FALSE, fig.height=2, fig.width=5}
ggplot(data=ToothGrowth, mapping=aes(x=factor(dose), y=len, color=supp)) + 
        geom_point() + theme_bw()+
        facet_wrap(~supp,ncol=2)+
        scale_color_manual(name="Vitamin C Sources",
                          labels=c("Orange Juice", "Ascorbic Acid"),
                          values=c("orange1", "yellow")) +
        theme(legend.key = element_rect(colour = NA), legend.position="bottom") +
        labs(title="Tooth Growth Absolute Values") + xlab("") +
        ylab("Tooth length, microns")
```

```{r summary 2, warning=FALSE, results='hide', echo=FALSE}
str(ToothGrowth)
ggplot(data=ToothGrowth, mapping=aes(x=factor(dose), y=len, fill=supp)) + 
        geom_boxplot() + 
        theme_bw()+
        labs(title="Dose and Source Dependence of the Tooth Growth") +
        theme(legend.key = element_rect(colour = NA), legend.position="bottom") +
        xlab("Source Dose, milligrams") +ylab("Tooth length, microns") +
        scale_fill_manual(name="Vitamin C Sources",
                            labels=c("Orange Juice", "Ascorbic Acid"),
                            values=c("orange1", "yellow"))
```
There obviously exists a positive and proportionate effect of orange juice as compared with the ascorbic acid, however, this effect can only be observed at dose levels below 2 mg. We can also demonstrate this in numeric values (see below, see **[Code Block 1][]**)
```{r summary 3}
ToothGrowth %>% filter(supp=="OJ") -> oj.data
ToothGrowth %>% filter(supp=="VC") -> vc.data

```
#References

[1]:http://jn.nutrition.org/content/33/5/491.full.pdf "C. I. Bliss (1952) The Statistics of Bioassay. Academic Press." 

#Related R Code

##Code Block 1
```{r code 0, results="hide",fig.keep='none', warning=FALSE}

```
