---
title: "ToothGrowth Data Exploratory Analysis"
author: "Alexander Tuzhikov"
date: "September 14, 2015"
output:
  pdf_document:
    fig_height: 5
    fig_width: 8
    highlight: haddock
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 1
  html_document:
    fig_caption: yes
    fig_height: 9
    fig_width: 12
    highlight: haddock
    keep_md: yes
    number_sections: yes
    toc: yes
mainfont: Ubuntu
documentclass: article
---
#Synopsis

We are now moving to the part 2 of the task. Below we will exlore the `ToothGrowth` dataset from R `datasets` package. The headers below will correspond to the tasks. The data comes from the study ["The Effect of Vitamin C on Tooth Growth in Guinea Pigs"][1]. The data is described as: *The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).*

#Load the ToothGrowth data and perform some basic exploratory data analyses

```{r libraries, message=FALSE, warning=FALSE, echo=FALSE}
library(datasets)
library(gridExtra)
library(ggplot2)
library(dplyr)
library(pander)
```

```{r data, message=FALSE, warning=FALSE}
data("ToothGrowth")
```

#Provide a basic summary of the data
Well, the data is presented as a `data.frame` of `r nrow(ToothGrowth)` samples in `r ncol(ToothGrowth)` rows: `r colnames(ToothGrowth)`;  The data is summarized in the Table 1 below (see **[Code Block 1][]**).

* **len** is the tooth length (microns)
* **supp** is the supplement type (VC or OJ)
* **dose** is the dose (milligrams)

```{r summary 1, echo=FALSE, message=FALSE}                         
ToothGrowth %>% mutate(dose=factor(dose)) %>% 
        group_by(supp, dose) %>% 
        summarize(
                min.len=min(len), 
                max.len=max(len),
                mean.len=mean(len),
                sd.len=sd(len),
                q25=quantile(len,probs =0.25),
                median.len=median(len),
                q75=quantile(len,probs =0.75)) %>%
        pander(caption = "Summary of the \"ToothGrowth\" Data")
```

```{r summary 2, warning=FALSE, results='hide', echo=FALSE}
str(ToothGrowth)
ggplot(data=ToothGrowth, mapping=aes(x=factor(dose), y=len, fill=supp)) + 
        geom_boxplot() + 
        theme_bw()+
        labs(title="Dose and Source Dependence of the Tooth Growth") +
        theme(legend.key = element_rect(colour = NA), legend.position="bottom") +
        xlab("Source Dose, milligrams") +ylab("Tooth length, microns") +
        scale_fill_manual(name="Vitamin C Sources",
                            labels=c("Orange Juice", "Ascorbic Acid"),
                            values=c("orange1", "yellow"))
```

There obviously exists a positive and proportionate effect of orange juice as compared with the ascorbic acid, however, this effect can only be observed at dose levels below 2 mg. We can also demonstrate this in numeric values (see below, see **[Code Block 2][]**)

```{r summary 3, echo=FALSE}
tooth.grows.list<- list()
ToothGrowth %>% 
        group_by(supp, dose) %>% 
        do({
                
                tooth.grows.list[[length(tooth.grows.list)+1]]<<- as.data.frame(.)
                names(tooth.grows.list)[length(tooth.grows.list)]<<-paste(.$supp[1],.$dose[1],sep="_")
                return(.)
        }) -> ToothGrowth
pander(t.test(tooth.grows.list[["OJ_0.5"]]$len, tooth.grows.list[["VC_0.5"]]$len), caption = "Orange Juice vs Ascorbic Acid at 0.5 mg Dose")
pander(t.test(tooth.grows.list[["OJ_1"]]$len, tooth.grows.list[["VC_1"]]$len), caption = "Orange Juice vs Ascorbic Acid at 1.0 mg Dose")
pander(t.test(tooth.grows.list[["OJ_2"]]$len, tooth.grows.list[["VC_2"]]$len), caption = "Orange Juice vs Ascorbic Acid at 2.0 mg Dose")
```

#References

[1]:http://jn.nutrition.org/content/33/5/491.full.pdf "C. I. Bliss (1952) The Statistics of Bioassay. Academic Press." 

#Related R Code

##Code Block 0
```{r code 0, results="hide",fig.keep='none', warning=FALSE}

```

##Code Block 1
```{r code 1, results="hide",fig.keep='none', warning=FALSE}

```

##Code Block 2
```{r code 2, results="hide",fig.keep='none', warning=FALSE}

```
